Increase of GDP and Monetary Expansion

### 2023-02-22 outputs/paper/paper.qmd
Final Revision

---
title: Increase of GDP and Monetary Expansion
subtitle: An Analysis of Potential Factors from 16th to 19th century
author: 
  - Rae Zhang
  - Jenny Shen
  - Sarah Zhang
thanks: "Code, data and reproduction package are available at: https://github.com/JunweiZhang130/Replication-for-The-real-effects-of-monetary-expansions-evidence-from-a-large-scale-history.git and https://www.socialsciencereproduction.org/reproductions/8d4a8d6c-9f9f-44dc-bd97-9c810ed677f6/index "
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper discusses the causal relationship between GDP growth and monetary expansion from the 16th to the 19th century, with the intent to identify the potential factors that contributed to economic growth during this period. Using a historical analysis approach, the paper examines multiple datasets including the nominal and real GDP of different European countries, the production of precious metals in the Americas, and the mint output of England that gave a glimpse to what potentially stimulates economic growth. The findings suggest that monetary injection played a significant role in increasing mint output and driving GDP growth during this period. Additionally, the study highlights the importance of monetary policies in promoting economic development. Overall, this paper provides insights into the factors that drove economic growth during a critical period in economic history. "
format: pdf
toc: FALSE
output: bookdown::pdf-document2
bibliography: references.bib
---

@@

\newpage
# 1 Introduction

  During the early period, the discovery of massive deposits of precious metals in the Americas resulted in monetary expansions to Europe’s supply. The finding of vast quantities of gold and silver in the New World had a significant influence on the world economy's subsequent history. The increase in the supply has reshaped the imports of consumer goods across Europe, and one of the results is monetary expansions. The effects of monetary policy have had major consequences, including increased economic activity, higher investment rates, and enhanced innovation. This paper explores the causal relationship of monetary policy and production of metals, and its effect on the increase of GDP, using historical data of the six European countries (citation). Specifically, we will examine how the real GDP of European countries, Gold and Silver production, and Mint output has impacted monetary expansions. Some factors that we considered which influenced monetary expansions on a variety of outcomes, including economic activity, investment, and innovation.

@@

  The analysis found that precious metals are one of the factors that contributed to monetary expansion in the 18th century. When precious metals, such as gold and silver were used as the basis for currency, it caused an increase in demand within the country, or within international trade, which led to monetary expansion. During the period of 1500 to 1800, production of precious metal increased which contributed to many of the European country’s GDP. In our reproduction of their paper, we build on their “monetary expansion” explanation by expanding on the relationship between Nominal GDP and Real GDP and the effect of precious metals in the period of 1500 to 1800.

@@

The datasets for this report were obtained from the data section "The Real Effects of Monetary Expansion: Evidence from a Large-scale Historical Experiment" [@PalmaRealEffect], a paper from the Review of Economic Studies. The data used by the paper was compiled by various sources that spans a very long horizon (1531–1790), The most common primary data sources for these studies are surviving account books of the landed estates formerly belonging to local government and royal administration, historical hospitals, prisons, charities, orphanages, universities, and institutions of the church, particularly monasteries and convents. GDP and prices come from the following sources: England from Broadberry et al. [@Broadberry2015], Holland from van Zanden and van Leeuwen [@vanZanden2012], Spain from ¡lvarez-Nogal and Prados de la Escosura [@AlvarezNogal2013], Italy from Malanima [@Malanima2011], Portugal from Palma and Reis (2019) and Germany from Pfister [@Pfister2020]. When the original price level was in silver it was transformed into monetary units using the information given by Karaman et al. [@Karaman2020]. Englandís nominal mint output from Challis[@Challis1992] and Palma's "Reconstruction of money supply over the long run: the case of England" and "Money and modernization in early modern England"[@Palma2018b]. With regards to the innovation component of gold arrivals, it is built using Costa et al. [@Costa2013] and Morineau[@Morineau2009] as explained in the text. The data is mainly numeric, counting the number of the amount of monetary activities in European countries.

The analysis will be carried out using the statistical programming language R [@citeR], using the haven [@citehaven] and tidyverse [@citetidyverse] packages. All figures in the report are generated using ggplot2 [@citeggplot2].

@@

# Loading in datasets and library
liquidity <- read_dta(here::here("scripts/liquidity.dta"))
datadescriptive <- read_dta(here::here("scripts/datadescriptive.dta"))

```
\newpage
# 3 Results

@@

NominalGDP <- liquidity |>
  ggplot(aes(x = year, 
             y = Nominal_GDPindex1700100, 
             color = country)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = color) +
  scale_y_continuous(limits = c(0, 300), 
                     expand = c(0, 0)) +
  labs(x = "Year", 
       y = "Nominal GDP, index(1700=100)", 
       title = "Nominal GDP for Six Countries during 1500 to 1800") +
  theme_minimal() +
  facet_wrap(~ country, ncol = 2)

@@

  labs(x = "Year", 
       y = "Real GDP, index(1700=100)", 
       title = "Real GDP for Six Countries during 1500 to 1800") +
  theme_minimal() +
  facet_wrap(~ country, ncol = 2)

@@

TotalGoldShare <- datadescriptive |>
  ggplot(aes(x = year, y = goldshare, fill = "Total Gold Share")) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  geom_line(aes(x = year, y = goldshare, color = "Total Gold Share"), size = 1.5) +
  labs(x = "Year", 
       y = "Total Gold Share", 
       title = "Share of Gold in Total Production through 1530 to 1790") +
  scale_x_continuous(limits = c(1530, 1790), breaks = seq(1530, 1790, by = 10)) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = mypal, guide = "none") +
  scale_color_manual(values = mypal, guide = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

TotalGoldShare
```
\newpage

@@

EnglandMintOutput <- datadescriptive %>%
  ggplot(aes(x = year, y = englandmintmillions)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  labs(x = "Year", 
       y = "England Mint Output (in millions)", 
       title = "England Mint Output from 1531 to 1790") +
  scale_x_continuous(limits = c(1531, 1790))
EnglandMintOutput
```

@@

## Cohort Effect in European Countries

@@

## A Case of Monetary Expansion: England’s Mint Output
[@fig-mintoutput] illustrates the English nominal mint output from 1531 - 1790 period. Mint output refers to the amount of new currency that is produced over the period of time, which can be measured using the quantity of the coins that are produced. The Mint output was a steady decrease in the issue of gold coins and sharper increase in the production of silver coins. This is primarily caused by the proportion of silver to gold coins. The mint output was relatively low in the fourteenth century, but as the picture indicates, it surged from the sixteenth century onward. During the 16th century, the world gold-silver ratio grew, as world stocks increased, the less precious metal continued to depreciate.

@@

## How Does Mint Output and Monetary Expansion Contribute to Today’s Economy?

Mint output impacts today’s economy in several ways including facilitating transactions, generating revenue, supporting financial stability, and promoting national identity. In daily life, increasing mint output makes it easier and more convenient for people to purchase goods and services, pay bills, and conduct financial transactions. While the process of producing new currencies and coins helps generate revenue for the government, which can then be used to fund public services and infrastructures. A reliable and consistent mint output could also make certain an adequate amount of currency in circulation, ensuring there will not be disruptions in the financial system. Furthermore, currencies and coins often feature national symbols and images that promote a country’s cultural identity and heritage, creating a positive and lasting effect on tourism and international trade.

On the other hand, the effects of monetary expansion on economic activities are more complex. Monetary expansion can affect the economy through changes in interest rates, money supply,  exchange rates, and GDP. The decrease in interest rates could stimulate economic growth by encouraging individuals and businesses to borrow more money in order to spend and make investments. However, it can also lead to inflation when the amount of mint output becomes greater than the supply of goods and services in the economy. Additionally, monetary expansion can depreciate the value of the currency since a lower interest rate often makes a country’s currency less attractive to investors from a foreign country. And from the findings of this paper, monetary injections are proven to be the cause of increase in both nominal and real GDP of different countries. Overall, monetary expansion can have positive effects such as stimulating the growth of the economy, but it can also instigate negative effects such as inflation if not managed properly.

## Bias and Ethical Concerns

@@

# 5 Weakness
In the reproduction process, we translate the original STATA language into R, meanwhile,  we updated the profile name to make it better organized in the folder. However, the paper still has many weaknesses. For example, the information Palma provided in the package was incomplete. Additionally, considering source was collected from local governments and account books, and the data was from an old range of period. Errors in the dataset can occur. Due to this, the conclusion that we draw from this graph are likely not accurate to reflect the overall trend of GDP. Our research question focused on the trend of GDP over the period of 1500 to 1800, and only focusing on the precious metal. Yet our conclusion was drawn from the six countries and does not reflect the world.

>>>>

---
title: Increase of GDP and Monetary Expansion
subtitle: An Analysis of Potential Factors from 16th to 19th century
author: 
  - Rae Zhang
  - Jenny Shen
  - Sarah Zhang
thanks: "Code, data and reproduction package are available at: https://github.com/JunweiZhang130/Replication-for-The-real-effects-of-monetary-expansions-evidence-from-a-large-scale-history.git and https://www.socialsciencereproduction.org/reproductions/8d4a8d6c-9f9f-44dc-bd97-9c810ed677f6/index "
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper discusses the causal relationship between GDP growth and monetary expansion from the 16th to the 19th century, with the intent to identify the potential factors that contributed to economic growth during this period. Using a historical analysis approach, the paper examines multiple datasets including the nominal and real GDP of different European countries, the production of precious metals in the Americas, and the mint output of England that gave a glimpse to what potentially stimulates economic growth. The findings suggest that monetary injection played a significant role in increasing mint output and driving GDP growth during this period. Additionally, the study highlights the importance of monetary policies in promoting economic development. Overall, this paper provides insights into the factors that drove economic growth during a critical period in economic history. "
format: pdf
output: bookdown::pdf-document2
bibliography: references.bib
---

@@

# 1 Introduction

During the early period, the discovery of massive deposits of precious metals in the Americas resulted in monetary expansions to Europe's supply. The finding of vast quantities of gold and silver in the New World had a significant influence on the world economy's subsequent history. The increase in the supply has reshaped the imports of consumer goods across Europe, and one of the results is monetary expansions. The effects of monetary policy have had major consequences, including increased economic activity, higher investment rates, and enhanced innovation. This paper explores the causal relationship of monetary policy and production of metals, and its effect on the increase of GDP, using historical data of the six European countries [@PalmaRealEffect]. Specifically, we will examine how the real GDP of European countries, Gold and Silver production, and Mint output has impacted monetary expansions. Some factors that we considered which influenced monetary expansions on a variety of outcomes, including economic activity, investment, and innovation.

@@

The analysis found that precious metals are one of the factors that contributed to monetary expansion in the 18th century. When precious metals, such as gold and silver were used as the basis for currency, it caused an increase in demand within the country, or within international trade, which led to monetary expansion. During the period of 1500 to 1800, production of precious metal increased which contributed to many of the European country's GDP. In our reproduction of their paper, we build on their "monetary expansion" explanation by expanding on the relationship between Nominal GDP and Real GDP and the effect of precious metals in the period of 1500 to 1800.

@@

The datasets for this report were obtained from the data section "The Real Effects of Monetary Expansion: Evidence from a Large-scale Historical Experiment" [@PalmaRealEffect], a paper from the Review of Economic Studies. The data used by the paper was compiled by various sources that spans a very long horizon (1531--1790), The most common primary data sources for these studies are surviving account books of the landed estates formerly belonging to local government and royal administration, historical hospitals, prisons, charities, orphanages, universities, and institutions of the church, particularly monasteries and convents. GDP and prices come from the following sources: England from Broadberry et al. [@Broadberry2015], Holland from van Zanden and van Leeuwen [@vanZanden2012], Spain from ¡lvarez-Nogal and Prados de la Escosura [@AlvarezNogal2013], Italy from Malanima [@Malanima2011], Portugal from Palma and Reis (2019) and Germany from Pfister [@Pfister2020]. When the original price level was in silver it was transformed into monetary units using the information given by Karaman et al. [@Karaman2020]. Englandís nominal mint output from Challis[@Challis1992] and Palma's "Reconstruction of money supply over the long run: the case of England" and "Money and modernization in early modern England"[@Palma2018b]. With regards to the innovation component of gold arrivals, it is built using Costa et al. [@Costa2013] and Morineau[@Morineau2009] as explained in the text. The data is mainly numeric, counting the number of the amount of monetary activities in European countries.

The analysis will be carried out using the statistical programming language `R` [@citeR], using the `haven` [@citehaven] and `tidyverse` [@citetidyverse] packages. All figures in the report are generated using `ggplot2` [@citeggplot2].

@@

# Loading in datasets and library

liquidity <- read_dta(here::here("scripts/liquidity.dta"))
datadescriptive <- read_dta(here::here("scripts/datadescriptive.dta"))

# 3 Results

@@

NominalGDP <- liquidity |>
  ggplot(aes(x = year, 
             y = Nominal_GDPindex1700100, 
             color = country)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = color) +
  scale_y_continuous(limits = c(0, 300), 
                     expand = c(0, 0)) +
  labs(x = "Year", 
       y = "Nominal GDP, index(1700=100)", 
       title = "Nominal GDP for Six Countries during 1500 to 1800") +
  theme_minimal() +
  facet_wrap(~ country, ncol = 2)+
  theme(plot.title = element_text(hjust = 0.5))

@@

  labs(x = "Year", 
       y = "Real GDP, index(1700=100)", 
       title = "Real GDP for Six Countries during 1500 to 1800") +
  theme_minimal() +
  facet_wrap(~ country, ncol = 2) +
  theme(plot.title = element_text(hjust = 0.5))

@@

TotalGoldShare <- datadescriptive |>
  ggplot(aes(x = year, y = goldshare, fill = "Total Gold Share")) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  geom_line(aes(x = year, y = goldshare, color = "Total Gold Share"), size = 1.5) +
  labs(x = "Year", 
       y = "Total Gold Share", 
       title = "Share of Gold in Total Production through 1530 to 1790") +
  scale_x_continuous(limits = c(1530, 1790), breaks = seq(1530, 1790, by = 10)) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = mypal, guide = "none") +
  scale_color_manual(values = mypal, guide = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  theme(plot.title = element_text(hjust = 0.5))

TotalGoldShare
```

@@

EnglandMintOutput <- datadescriptive %>%
  ggplot(aes(x = year, y = englandmintmillions)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  labs(x = "Year", 
       y = "England Mint Output (in millions of £)", 
       title = "England Mint Output from 1531 to 1790") +
  scale_x_continuous(limits = c(1531, 1790))+
  # Adjust plot title alignment
  theme(plot.title = element_text(hjust = 0.5))
EnglandMintOutput
```

@@

## 4.1 Cohort Effect in European Countries

@@

## 4.2 A Case of Monetary Expansion: England's Mint Output

[@fig-mintoutput] illustrates the English nominal mint output from 1531 - 1790 period. Mint output refers to the amount of new currency that is produced over the period of time, which can be measured using the quantity of the coins that are produced. The Mint output was a steady decrease in the issue of gold coins and sharper increase in the production of silver coins. This is primarily caused by the proportion of silver to gold coins. The mint output was relatively low in the fourteenth century, but as the picture indicates, it surged from the sixteenth century onward. During the 16th century, the world gold-silver ratio grew, as world stocks increased, the less precious metal continued to depreciate.

@@

## 4.3 How Does Mint Output and Monetary Expansion Contribute to Today's Economy?

Mint output impacts today's economy in several ways including facilitating transactions, generating revenue, supporting financial stability, and promoting national identity. In daily life, increasing mint output makes it easier and more convenient for people to purchase goods and services, pay bills, and conduct financial transactions. While the process of producing new currencies and coins helps generate revenue for the government, which can then be used to fund public services and infrastructures. A reliable and consistent mint output could also make certain an adequate amount of currency in circulation, ensuring there will not be disruptions in the financial system. Furthermore, currencies and coins often feature national symbols and images that promote a country's cultural identity and heritage, creating a positive and lasting effect on tourism and international trade.

On the other hand, the effects of monetary expansion on economic activities are more complex. Monetary expansion can affect the economy through changes in interest rates, money supply, exchange rates, and GDP. The decrease in interest rates could stimulate economic growth by encouraging individuals and businesses to borrow more money in order to spend and make investments. However, it can also lead to inflation when the amount of mint output becomes greater than the supply of goods and services in the economy. Additionally, monetary expansion can depreciate the value of the currency since a lower interest rate often makes a country's currency less attractive to investors from a foreign country. And from the findings of this paper, monetary injections are proven to be the cause of increase in both nominal and real GDP of different countries. Overall, monetary expansion can have positive effects such as stimulating the growth of the economy, but it can also instigate negative effects such as inflation if not managed properly.

## 4.4 Bias and Ethical Concerns

@@

# 5 Weakness

In the reproduction process, we translate the original STATA language into R, meanwhile, we updated the profile name to make it better organized in the folder. However, the paper still has many weaknesses. For example, the information Palma provided in the package was incomplete. Additionally, considering source was collected from local governments and account books, and the data was from an old range of period. Errors in the dataset can occur. Due to this, the conclusion that we draw from this graph are likely not accurate to reflect the overall trend of GDP. Our research question focused on the trend of GDP over the period of 1500 to 1800, and only focusing on the precious metal. Yet our conclusion was drawn from the six countries and does not reflect the world.
